A Kleisli-style construction for Parametric Monads, with Examples

### 2018-11-16 effects/effects.tex
Finished off the paper.

% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{xspace}
\usepackage{IEEEtrantools}
\usepackage{mathtools}
\usetikzlibrary{cd}
\usepackage{mathpartir}
\def\UrlFont{\rmfamily}

\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
    \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}

@@

@@

\newcommand{\coin}{\textsf{coin}}
\newcommand{\polydie}{\textsf{polydie}}
\newcommand{\bool}{\textsf{bool}}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{A Kleisli-like construction for Parametric Monads, with Examples}
%
\titlerunning{Dum-di-dum}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{}
%
\authorrunning{} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{}
%
\institute{University of Bath, Claverton Down Road, Bath.  
BA2 7AY,\\
\email{wjg27@bath.ac.uk},\\ WWW home page:
\texttt{http://people.bath.ac.uk/wjg27}}

@@

By cutting down the category $\DProb$ we can obtain slightly different notions of probability.  
For our purposes, we will be using the full subcategory $\BProb$ where the objects $(X,\bP_X)$ are those that satisfy the following three additional conditions.
\begin{description}
  \item[Countability] The set $X$ is countable.
  \item[Finite support] There is a finite subset $E\subseteq X$ such that $\bP_X(E)=1$.
  \item[Rational probabilities] $\bP_X(A)\in\mathbb{Q}$ for all $A\subseteq X$.
\end{description}

The action of $\BProb$ gives rise to a monad on the category of countable sets.

@@

\cite{DanosHarmer} gives a fully abstract game semantics for the language Probabilistic Algol (PA).  
The base language for PA is the language Idealized Algol (IA) from \cite{SamsonGuyIAPassive}, and it has been extended with a constant $\coin\from\bool$.  
We give the language a small-step reduction semantics, as shown in Figure \ref{pa-operational}.

\begin{figure*}
  \begin{mathpar}
    \inferrule*{ }{\ia s {(\lambda x.M)\;N} \opto \ia s {M[N/x]}}
    \and
    \inferrule*{ }{\ia s {\Y_T M}\opto \ia s {M (\Y_T M)}}
    \and
    \inferrule*{ }{\ia s {\suc \n}\opto\ia s {\n + \mathsf{1}}}
    \and
    \inferrule*{ }{\ia s {\pred \n}\opto\ia s {\0 \sqcup (\n - \mathsf{1})}}
    \and
    \inferrule*{ }{\ia s {\IfO \0 M N} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\IfO (\n + \mathsf{1}) M N} \opto \ia s N}
    \and
    \inferrule*{ }{\ia s {\If \true\;M\;N} \opto M}
    \and
    \inferrule*{ }{\ia s {\If \false\;M\;N} \opto N}
    \and
    \inferrule*{ }{\ia s {\deref (\mkvar M N)} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {(\mkvar M N) \coloneqq L} \opto \ia s {N\;L}}
    \and
    \inferrule*{ }{\ia s {v \coloneqq n} \opto \ia {\stup s v n} \skipp}
    \and
    \inferrule*[left={$s(v)=n$}]{ }{\ia s {\deref v} \opto \ia s n}
    \and
    \inferrule*{ }{\ia s {\skipp;M} \opto \ia s M}
    \and
    \inferrule*{ }{\ia s {\new_T \lambda v.M}\opto \ia {\stup s v 0} M}
    \and
    \inferrule*{\ia s M \opto \ia s {M'}}{\ia s {E[M]} \opto \ia s {E[M']}}
    \and
    \inferrule*[left=coin]{ }{\ia s \coin \opto \ia s \true}
    \and
    \inferrule*[left=coin]{ }{\ia s \coin \opto \ia s \false}
  \end{mathpar}
  \caption{Small-step operational semantics for Probabilistic Algol, where $E$ ranges over suitable evaluation contexts}
  \label{pa-operational}
\end{figure*}

All of these rules are deterministic with the exception of the two marked \LeftTirName{coin}.
Given an evaluation $\pi = M \to M_1 \to \cdots \to \x$ of a term $M$ of ground type $o$, we define the \emph{weight} $w(\pi)$ of $\pi$ to be $\frac{1}{2^k}$, where $k$ is the number of instances of a \LeftTirName{coin} rule occurring in $\pi$.

This defines a discrete sub-probability measure on the set corresponding to the ground type $o$ given by
\[
  \bP(\{x\}) = \sum_{\substack{\text{evaluations}\\\pi = M \to\dots \x}}w(\pi)\,.
  \]
This is only a subprobability measure, because some evaluations may continue forever.

\cite{DanosHarmer} shows how we can define other probabilistic terms using $\coin$.  
Specifically, they define a term $\polydie\from\bN\to\bN$, where the behaviour of $\polydie n$ is to return the result of tossing an $n$-sided die (with sides numbered $0$ to $n-1$), with each outcome occurring with probability $1/n$.  
The definition of $\polydie n$ in terms of $\coin$ works as follows.  
We choose some number $k$ such that $2^k\ge n$ and then toss the coin $n$ times to give one of $2^k$ outcomes.  
If an outcome corresponds (under some fixed correspondence $n\hookrightarrow 2^k$) to a natural number below $n$, then we return that number; otherwise, we repeat the process until we get such a number.

@@

This gives rise to a reader-style action of $\BProb^{op}/\G$, and the resulting category $\G/\BProb^{op}$ is therefore a symmetric monoidal closed category.  
Unfortunately, the monoidal product on $\G/\BProb^{op}$ is not Cartesian; the problem is that there is no unique way to form the copairing of

\bibliographystyle{spbasic}
\bibliography{effects}

>>>>

% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{xspace}
\usepackage{IEEEtrantools}
\usepackage{mathtools}
\usetikzlibrary{cd}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\def\UrlFont{\rmfamily}

\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
    \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
\newcommand{\deno}[1]{\left\llbracket#1\right\rrbracket}
\renewcommand{\ts}{{\;\vdash}}

@@

\newcommand{\test}{\texttt{test}}

@@

\newcommand{\coin}{\textsf{coin}}
\newcommand{\polydie}{\textsf{polydie}}
\newcommand{\multicoin}{\textsf{multicoin}}
\newcommand{\bool}{\textsf{bool}}

@@

\begin{document}
\mainmatter              % start of a contribution
%
\title{A Kleisli-style construction for Parametric Monads, with Examples}
%
\titlerunning{A Kleisli-Style Construction}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{}
%
\authorrunning{} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{}
%
\institute{University of Bath, Claverton Down Road, Bath.  
BA2 7AY,\\
\email{wjg27@bath.ac.uk},\\ WWW home page:
\texttt{http://people.bath.ac.uk/wjg27}}

@@

By cutting down the category $\DProb$ we can obtain slightly different notions of probability.  
For our purposes, we will be using the full subcategory $\BProb$ where the objects $(X,\bP_X)$ are those that satisfy the following three additional conditions.
\begin{description}
  \item[Countability] The set $X$ is countable.
  \item[Finite support] There is a finite subset $E\subseteq X$ such that $\bP_X(E)=1$.
\end{description}

The action of $\BProb^{op}$ gives rise to a monad on the category of countable sets.

@@

\cite{DanosHarmer} gives a fully abstract game semantics for the language Probabilistic Algol (PA).  
The base language for PA is the language Idealized Algol (IA) from \cite{SamsonGuyIAPassive}, and it has been extended with a constants $\coin\from\bool$.  
To make things a little more interesting, we will extend this further to incorporate an infinite family of constants $\coin_p\from\bool$, where $p$ ranges over the real interval $[0,1]$.

We give the language a small-step operational semantics.  
The rules for the deterministic terms may be derived from the operational semantics given in \cite{SamsonGuyIAPassive}; for example:
\[
  \inferrule*{ }{\ia s {v \coloneqq n} \opto \ia {\stup s v n} \skipp}
  \]
See also \cite{JimGuySmallStep}.

We add two small-step rules for the $\coin_p$ primitives:

\begin{mathpar}
  \inferrule*[left={heads${}_p$}]{ }{\ia s \coin_p \opto \ia s \true}
  \and
  \inferrule*[left={tails${}_{1-p}$}]{ }{\ia s \coin_p \opto \ia s \false}\,.
\end{mathpar}

These are the only nondeterministic parts of the operational semantics.

Given a small-step evaluation $\pi = M \to M_1 \to \cdots \to \x$ of a term $M$ of ground type $o$, we define the \emph{probability} $p(\pi)$ of $\pi$ to be the product of all the indices $q$ belonging to \LeftTirName{heads${}_q$} or \LeftTirName{tails${}_q$} rules.

This defines a discrete sub-probability measure on the set corresponding to the ground type $o$ given by
\[
  \bP(\{x\}) = \sum_{\substack{\text{evaluations}\\\pi = M \to\dots \x}}p(\pi)\,.
  \]
This is only a subprobability measure, because some evaluations may continue forever.

\cite{DanosHarmer} shows how we can define other probabilistic terms using $\coin$.  
Specifically, they define a term $\polydie\from\bN\to\bN$, where the behaviour of $\polydie\,n$ is to return the result of tossing an $n$-sided die (with sides numbered $0$ to $n-1$), with each outcome occurring with probability $1/n$.  
The definition of $\polydie\,n$ in terms of $\coin$ works as follows.  
We choose some number $k$ such that $2^k\ge n$ and then toss the coin $n$ times to give one of $2^k$ outcomes.  
If an outcome corresponds (under some fixed correspondence $n\hookrightarrow 2^k$) to a natural number below $n$, then we return that number; otherwise, we repeat the process until we get such a number.

@@

This gives rise to a reader-style action of $\BProb^{op}/\G$, and the resulting category $\G/\BProb^{op}$ is therefore a symmetric monoidal closed category.

\subsection{Testing morphisms and plays}

The equivalence relation on morphisms is still too weak for our purposes.  
For example, if the boolean type $\bB$ carries a Bernoulli distribution with coefficient $p$, then the diagonal map $\Delta\from \bB\to \bB\times\bB$ should be equivalent to the identity $\id\from\bB\times\bB\to\bB\times\bB$ as terms of type $\bB\times\bB$, because the projections on to each component are equivalent.  
However, they are not yet equivalent in our category.

To improve the equivalence relation, we introduce the concept of \emph{testing morphism} for an object $A$ that is the denotation of an Idealized Algol term.
This is a term of type $(\bN \to A) \to A$ that is the denotation of an Idealized Algol term of the form
\[
  \test_{n_1,\cdots,n_k} = \lambda f.\new\;v\coloneqq 0 \text{ in } f(v \coloneqq \deref v + 1; \case\,v\,\_\,n_1\,\cdots\,n_k\,\Omega)\,,
  \]
where $n_1,\cdots,n_k$ is a sequence of numbers.  
The effect of this term is to call $f$, and to pass in the numbers $n_1,\cdots,n_k$ in turn, and diverging if $f$ calls its argument more than $k$ times.
In game semantics, the sequence $n_1,\cdots,n_k$ is more easily recognized as a sequence of moves occurring in the game $\bN$ on the left.

Given a morphism $\sigma\from \bN \to A$, we say that the sequence $n_1,\cdots,n_k$ is a \emph{play} of $\sigma$ if it is minimal (with respect to the prefix ordering) such that $\test_{n_1,\cdots,n_k}(\sigma)$ represents a non-diverging term.
In that case, the \emph{result} of the play is the strategy $\test_{n_1,\cdots,n_k}(\sigma)$ for $A$.

Now, if the set $\bN$ carries a probability distribution, then this gives rise to a sub-probability distribution on the set of plays for $\sigma$, where the probability associated to the sequence $n_1,\cdots,n_k$ is the product of the probabilities of the $n_i$.  
This then induces a discrete probability distribution on the set of morphisms $1\to A$.

We say that two morphisms $\sigma,\tau\from \bN\to A$ from $1$ to $A$ in $\G/\BProb$ are equivalent if they give rise to the same probability distribution on the set of strategies for $A$.  
This equivalence relation refines our earlier one.

\subsection{Single-threaded morphisms}

Our category is symmetric monoidal closed, but not yet Cartesian closed.  
A trick that is often used in Game Semantics (see, for example, \cite{mcCHFiniteND} and \cite{DanosHarmer}) is to pass to a category of \emph{single-threaded} strategies; this is equivalent to cutting down to those morphisms that are comonoid homomorphisms.

Note that, in our category, every object has a natural commutative comonoid structure that arises from diagonal $\Delta\from A\to A\times A$ in the original category.  
We say that a morphism $\sigma\from A\to B$ is \emph{single-threaded} if it is a homomorphism of the comonoid structures on $A$ and $B$; i.e., if the following square commutes.
\[
  \begin{tikzcd}
    A \arrow[r, "\sigma"] \arrow[d, "\Delta"']
      & B \arrow[d, "\Delta"] \\
    A \times A \arrow[r, "\sigma\times\sigma"]
      & B \times B
  \end{tikzcd}
  \]
Since the category of commutative comonoids in a symmetric monoidal category is automatically Cartesian closed (see, for example, \cite{Freyd}), this gives us a Cartesian closed category.

The deterministic terms are all comonoid homomorphisms.  
As long as we have applied the equivalence relation outlined above, the natural morphisms $\phi_{X,A}$ are also comonoid homomorphisms.

\subsection{Denotational semantics and adequacy}

We may now inductively build up a denotational semantics for PA.  
The deterministic terms may all be defined inside the base category $\G$ and transported into our new category via the natural inclusion.  
The probabilistic term $\coin_p$ is given by the natural morphism $\phi_{B_p, 1}$, where $B_p$ is a Bernoulli space with coefficient $p$.  
That is to say, it is given by the pair
\[
  ((\{H,T\}, \mu(H)=p, \mu(T)=1-p), 1 \to (\{H,T\} \to \bB))\,
  \]
where the function on the right sends $H$ to $\true$ and $T$ to $\false$.

Note now that the denotation of a term $M = C[\coin_{p_1},\cdots,\coin_{p_j}]T$ is given by the following morphism.
\[
  \bB^j
  \xrightarrow{\deno{b_1,\cdots,b_j\ts C[b_1,\cdots,b_j]}}
  \deno{T}\,,
  \]
where the distribution on the product $\bB^j$ is the product of Bernoulli distributions.

Fix an injection $r\from\bB^j\to \bN$ (for instance, the one given by ibinary representation) and suitable projection functions $\pr_i \from \bN \to \bB$ such that $\pr_i(r(v))$ is the $i$-th entry of $v$.
Let $\multicoin\from\nat$ be the term (depending on the values of the $p_i$) given by
\[
  (r\;\coin_{p_1}\,\cdots\,\coin_{p_j})\,.
  \]
Now the operational semantics of $\pr_i(\multicoin)$ is the same as the operational semantics of $\coin_{p_i}$, since the projections from the product are probability-preserving.  
This means that any evaluation $\pi$ of the term $M$ may be pegged to an evaluation of the term formed by replacing each instance of $\coin_{p_i}$ in $M$ with $\pr_i(\multicoin)$.

\begin{proposition}[Computational Adequacy]
  Let $M = C[\multicoin]\from \com$ be a term of Idealized Algol, where $C$ is a multi-holed deterministic IA context.
  Let $p$ be the sum of the probabilities of all evaluation sequences of $M$ that terminate at $\skipp$.

  The denotation of $M$ is a morphism $\bN \to \bC$ as outlined above.

  Then we get an induced sub-probability distribution on the set of morphisms $1\to A$.  
  The adequacy result is that $p$ is the same as the induced probability of the morphism $\deno{\skipp}$.
\end{proposition}
\begin{proof}
  Each terminating evaluation sequence of $M$ corresponds to a finite sequence $n_1,\cdots,n_k$ of the different values that $\multicoin$ takes on over the course of the evaluation of the term.  
  By inspection, we may peg this evaluation of $M$ exactly to the evaluation of the term $\test_{n_1,\cdots,n_k} (\lambda n.C[n])$ in pure deterministic Idealized Algol, and therefore this IA term must converge.  
  By the Adequacy result for Idealized Algol \cite{SamsonGuyIAPassive}, an IA term converges to $\skipp$ if and only if its denotation is $\deno{\skipp}$.  
  Therefore, there is bijective matching between terminating evaluation sequences of $M$ and plays of the morphism $\deno{M}$, with each one getting the same probability.  
  The result follows.
\end{proof}

\subsection{Full Abstraction}

In order to prove full abstraction for our model, we will define a further equivalence on morphisms.  
Say that two morphisms $f,g\from A \to B$ are \emph{intrinsically equivalent} if and only if for all morphisms $\alpha\from (A\to B)\to\bC$, $\alpha(f)$ and $\alpha(g)$ are probabilistically equivalent in the sense given above.
Then our full abstraction result becomes:

\begin{proposition}[Full Abstraction]
  Let $M,N\from T$ be two PA terms.  
  Then $M$ and $N$ are observationally equivalent if and only if their denotations $\deno{M}$ and $\deno{N}$ are intrinisically equivalent.
\end{proposition}

One direction (soundness) follows immediately from our Computational Adequacy result.  
To prove the other direction, it suffices to show that any compact strategy is definable \cite{CurienFullAbstraction}, where we say that a morphism $\sigma\from A \to (X \to B)$ from $A$ to $B$ is \emph{compact} if it is compact when considered as a morphism in $\G$.
We prove this using a factorization result:
\begin{lemma}[Deterministic factorization]
  Let $\sigma\from A \to B$ be a morphism in $\G/\BProb^{op}$.  
  Then $\sigma$ may be factorized as $\text{Det}(\sigma)(\multicoin)$, where $\text{Det}(\sigma)\from \bN \to A \to B$ is a strategy in the image of the inclusion functor $J\from \G \to \G/\BProb^{op}$, and $\multicoin$ is defined for some values $p_1,\cdots,p_k$.
\end{lemma}
\begin{proof}
  Treat $\sigma$ as a morphism $A \to X \to B$ in $\G$.  
  Since $X$ is finitely supported, there is a probability-preserving function $\bN \to X$, for some finitely supported distribution on $\bN$.  
  So $\sigma$ is equivalent to a morphism $A \to \bN \to B$, and then the factorization is automatic.
\end{proof}

\bibliographystyle{spphys}
\bibliography{effects}
